---
title: "Patient Medication Adherence Prediction"
output: html_notebook
---

The patient medication adherence classification project is a critical endeavor in healthcare aimed at predicting and categorizing patients' adherence to prescribed medications. Leveraging a dataset that includes variables such as patient age, prescription duration, medical conditions like diabetes and hypertension, lifestyle factors like smoking and alcoholism, and SMS reminders, the project seeks to develop a robust model for patient classification. By accurately distinguishing between adherent and non-adherent patients, healthcare providers can tailor interventions and support to enhance medication compliance, ultimately improving patient outcomes and reducing healthcare costs. This project represents a significant step toward more effective patient-centered care and better healthcare management. 
<br/><br/>
<h1><b>Loading dataset</b></h1>
<br/><br/>
<h3>
Loading dataset path</h3>
```{r}
file_path <- "D:/College/Assignments/Semester 5/Theory/Introduction to Data Science/dataset.csv"
```

<br/><br/>
<h3>
Loading dataset</h3>

```{r}
data <- read.csv(file_path)
```
<br/><br/>
<h1><b>Exploratory Data Analysis</b></h1>
<br/><br/>
<h3>
Analyzing Dimensions of data</h3>
```{r}
dim(data)
```
The dataset has 180212 rows and 11 columns
<br/><br/>
<h3>
Visualizing first few rows of data</h3>

```{r}
head(data)
```
Adherence is the target variable. Others columns are thefeatures.
<br/><br/>
<h3>
Visualizing structure of data as R object</h3>
```{r}
str(data)
```
<br/><br/>
<h3>
Summary structure of data</h3>
```{r}
summary(data)
```
<br/><br/>
<h3>
Calculating number of null values in data</h3>
```{r}
colSums(is.na(data))
```
The data has no null values.
<br/><br/>
<h3>
Loading psych library</h3>
```{r}
library(psych)
```
<br/><br/>
<h3>
Description of data</h3>
```{r}
describe(data)
```
<br/><br/>
<h1><b>Data Visualization</b></h1>
<br/><br/>
<h3>Visualizing category wise frequency of target variable Adherence</h3>
```{r}
# Calculate the category wise frequency of Adherence values
temp <- table(data$Adherence)
temp

# Calculate the relative frequencies
relative_frequencies <- round(prop.table(temp), 2)

# Create a bar plot
barplot(relative_frequencies, 
        names.arg = names(temp),
        col = "blue",
        main = "Adherence Distribution",
        xlab = "Adherence Values",
        ylab = "Relative Frequency")


```

The target class exhibits an imbalance, and sampling techniques will not be utilized. Given that the primary focus in the medical field is on precision and recall, accuracy is not the sole determinant of performance.
<br/><br/>
<h3>Visualizing category wise frequency of feature Gender</h3>

```{r}
# Print the size of each group
table(data$Gender)

# Create a count plot
barplot(table(data$Gender), names.arg = names(table(data$Gender)))
```
<br/><br/>
<h3>Visualizing category wise frequency of feature Diabetes</h3>
```{r}
# Print the size of each group
table(data$Diabetes)

# Create a count plot
barplot(table(data$Diabetes), names.arg = names(table(data$Diabetes)))

```
<br/><br/>
<h3>Visualizing category wise frequency of feature Alcoholism</h3>
```{r}
# Print the size of each group
table(data$Alcoholism)

# Create a count plot
barplot(table(data$Alcoholism), names.arg = names(table(data$Alcoholism)))

```
<br/><br/>
<h3>Visualizing category wise frequency of feature Alcoholism</h3>
```{r}
# Print the size of each group
table(data$HyperTension)

# Create a count plot
barplot(table(data$HyperTension), names.arg = names(table(data$HyperTension)))

```
<br/><br/>
<h3>Visualizing category wise frequency of feature Tuberculosis</h3>
```{r}
# Print the size of each group
table(data$Tuberculosis)

# Create a count plot
barplot(table(data$Tuberculosis), names.arg = names(table(data$Tuberculosis)))

```
<br/><br/>
<h3>Visualizing category wise frequency of feature Smokes</h3>
```{r}
# Print the size of each group
table(data$Smokes)

# Create a count plot
barplot(table(data$Smokes), names.arg = names(table(data$Smokes)))
```
<br/><br/>
<h1><b>Multivriate Data Visualization</b></h1>
<br><br/>
<h3>Genderwise Adherence</h3>
<br><br/>
<h4>Tabular description</h4>
```{r}
# Use table() to count combinations of Gender and Adherence
gender_adherence_table <- table(data$Gender, data$Adherence)

# Convert the table to a data frame
Gender_Adherence <- as.data.frame(gender_adherence_table)

# Rename the columns
colnames(Gender_Adherence) <- c("Gender", "Adherence", "Count")

# Display the resulting data frame
Gender_Adherence

```
<br><br/>
<h4>Visualization</h4>
```{r}
# Load the ggplot2 library
library(ggplot2)

# Create a ggplot object
ggplot(data = Gender_Adherence, aes(x = Gender, y = Count, fill = Adherence)) +
  geom_bar(stat = "identity") +
  labs(title = "Gender and Adherence") +
  theme_minimal()

```
<br><br/>
<h3>Tuberculosis and Adherence</h3>
<br><br/>
<h4>Tabular description</h4>
```{r}
# Create a table of counts
tuberculosis_adherence_table <- table(data$Tuberculosis, data$Adherence)

# Print the table
print(tuberculosis_adherence_table)

```
<br><br/>
<h4>Visualization</h4>
```{r}
tuberculosis_adherence_df <- as.data.frame(tuberculosis_adherence_table)

# Rename the columns
colnames(tuberculosis_adherence_df) <- c("Tuberculosis", "Adherence", "Count")

# Create a bar plot
ggplot(data = tuberculosis_adherence_df, aes(x = Tuberculosis, y = Count, fill = Adherence)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Tuberculosis and Adherence") +
  theme_minimal()

```
<br/><br/>
<h3>Smokes and Adherence</h3>
<br><br/>
<h4>Tabular description<h4>
```{r}
smokes_adherence_table <- table(data$Smokes, data$Adherence)
print(smokes_adherence_table)
```
<br><br/>
<h4>Visualization<h4>
```{r}
smokes_adherence_df <- as.data.frame(smokes_adherence_table)

# Rename the columns
colnames(smokes_adherence_df) <- c("Smokes", "Adherence", "Count")

# Create a grouped bar plot
ggplot(data = smokes_adherence_df, aes(x = Smokes, y = Count, fill = Adherence)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Smokes and Adherence") +
  theme_minimal()

```
<br><br/>
<h3>Hypertension and Adherence</h3>
<br><br/>
<h4>Tabular description<h4>
```{r}
# Create a table of counts
hypertension_adherence_table <- table(data$HyperTension, data$Adherence)
hypertension_adherence_table
```
<br><br/>
<h4>Visualization<h4>
```{r}
# Create a data frame from the table
hypertension_adherence_df <- as.data.frame(hypertension_adherence_table)

# Rename the columns
colnames(hypertension_adherence_df) <- c("HyperTension", "Adherence", "Count")

# Create a count plot with a hue
ggplot(data = hypertension_adherence_df, aes(x = HyperTension, y = Count, fill = Adherence)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "HyperTension and Adherence") +
  theme_minimal()
```
<br><br/>
<h3>Alcoholism and Adherence</h3>
<br><br/>
<h4>Tabular description<h4>
```{r}
alcoholism_adherence_table <- table(data$Alcoholism, data$Adherence)
alcoholism_adherence_table
```
<br><br/>
<h4>Visualization<h4>
```{r}
# Create a data frame from the table
alcoholism_adherence_df <- as.data.frame(alcoholism_adherence_table)

# Rename the columns
colnames(alcoholism_adherence_df) <- c("Alcoholism", "Adherence", "Count")

# Create a count plot with a hue
ggplot(data = alcoholism_adherence_df, aes(x = Alcoholism, y = Count, fill = Adherence)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Alcoholism and Adherence") +
  theme_minimal()
```
<br><br/>
<h3>Diabetes and Adherence</h3>
<br><br/>
<h4>Tabular description<h4>
```{r}
# Create a table of counts
diabetes_adherence_table <- table(data$Diabetes, data$Adherence)
diabetes_adherence_table
```
<br><br/>
<h4>Visualization<h4>
```{r}
# Create a data frame from the table
diabetes_adherence_df <- as.data.frame(diabetes_adherence_table)

# Rename the columns
colnames(diabetes_adherence_df) <- c("Diabetes", "Adherence", "Count")

# Create a count plot with a hue
ggplot(data = diabetes_adherence_df, aes(x = Diabetes, y = Count, fill = Adherence)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Diabetes and Adherence") +
  theme_minimal()

```
<br/><br/>
<h1><b>Data Pre-Processing</b></h1>
<br><br/>
<h3>Convert the Adherence and Gender column to integer 0 and 1.</h3>
<h4>Adherence Yes = 1 and No = 0, Gender M = 1, F = 0</h4>
```{r}
data$Adherence <- ifelse(data$Adherence == 'Yes', 1, 0)
data$Gender <- ifelse(data$Gender == 'M', 1, 0)
```
<br/><br/>
<h3>First few rows of data.</h3>
```{r}
head(data)
```
<br/><br/>
<h3>Setting patient_id as index of data frame.</h3>
```{r}
rownames(data) <- data$patient_id
```
<h3>First few rows of data.</h3>
```{r}
head(data)
```
<br/><br/>
<h3>Drop Column patientid id</h3>
```{r}
# Load the dplyr library
library(dplyr)

# Assuming 'data' is your data frame
data <- data %>% select(-patient_id)

```
<br/><br/>
<h3>First few rows of data.</h3>
```{r}
head(data)
```
<br/><br/>
<h3>Diabetes and Adherence table.</h3>
```{r}
# Assuming 'data' is your data frame
result_table <- table(data$Diabetes, data$Adherence)

# Create a data frame from the table with column names
result_df <- as.data.frame(result_table)
colnames(result_df) <- c("Diabetes", "Adherence", "Count")

# Display the resulting data frame
print(result_df)
```
<br/><br/>
<h3>Diabetes, alcoholism and Adherence table.</h3>
```{r}
result_table <- table(data$Diabetes, data$Alcoholism, data$Adherence)

# Create a data frame from the table with column names
result_df <- as.data.frame(result_table)
colnames(result_df) <- c("Diabetes", "Alcoholism", "Adherence", "Count")

# Display the resulting data frame
print(result_df)
```
<br/><br/>
<h3>Adherence table size.</h3>
```{r}
# Assuming 'data' is your data frame
table(data$Adherence)
```
<br/><br/>
<h1></b>Train Test and Split the Data.</b></h1>
<br/><br/>
<h3>Splitting data</h3>
```{r}
# Load the caTools library
library(caTools)
set.seed(123)

# Create a split tag
sample.split(data$Adherence, SplitRatio = 0.7) -> split_tag

# Subset the data into training and testing sets
train <- subset(data, split_tag == TRUE)
test <- subset(data, split_tag == FALSE)
```
<br/><br/>
<h3>First few columns of train data</h3>
```{r}
head(train)
```
<br/><br/>
<h3>First few columns of test data</h3>
```{r}
head(test)
```
<br/><br/>
<h1></b>Training and Evaluation models.</b></h1>
<br/><br/>
<h2><b>Logistic Regression</b></h2>
<br/><br/>
<h3>Training model</h3>
```{r}
# Assuming 'data' is your data frame with the features and the 'Adherence' target variable

# Fit a logistic regression model
logistic_model <- glm(Adherence ~ Age + Gender + Prescription_period + Diabetes + Alcoholism + HyperTension + Smokes + Tuberculosis + Sms_Reminder, data = data, family = "binomial")

# Summary of the logistic regression model
summary(logistic_model)

```
<br/><br/>
<h3>Testing model</h3>
```{r}
# Make predictions on the test dataset
predictions <- predict(logistic_model, newdata = test, type = "response")

# Evaluate the model's performance
# You can use various evaluation metrics depending on your problem, e.g., accuracy, ROC curve, AUC, confusion matrix, etc.

# For example, to compute accuracy:
actual_labels <- test$Adherence
log_predicted_labels <- ifelse(predictions > 0.5, 1, 0)  # Threshold at 0.5

# Calculate accuracy
accuracy <- mean(actual_labels == log_predicted_labels)
cat("Accuracy:", accuracy, "\n")

```
<br/><br/>
<h3>Testing model</h3>

```{r}
# Make predictions on the test dataset
predictions <- predict(logistic_model, newdata = test, type = "response")

# Evaluate the model's performance
# You can use various evaluation metrics depending on your problem, e.g., accuracy, ROC curve, AUC, confusion matrix, etc.

# For example, to compute accuracy:
actual_labels <- test$Adherence
log_predicted_labels <- ifelse(predictions > 0.5, 1, 0)  # Threshold at 0.5

# Calculate accuracy
accuracy <- mean(actual_labels == log_predicted_labels)
cat("Accuracy:", accuracy, "\n")

```
<br/><br/>
<h3>Confusion Matrix</h3>
```{r}
# Calculate the confusion matrix and classification report
library(caret)
confusion_matrix <- confusionMatrix(as.factor(log_predicted_labels), as.factor(actual_labels))

# Print the classification report
print(confusion_matrix)
```
<br/><br/>
<h3>Evaluation Metric</h3>
```{r}
# Extract precision, recall, F1-score, and support from the confusion matrix
precision <- confusion_matrix$byClass["Pos Pred Value"]
recall <- confusion_matrix$byClass["Sensitivity"]
f1_score <- confusion_matrix$byClass["F1"]
support <- confusion_matrix$byClass["Support"]

# Print the metrics
cat("Precision:", precision, "\n")
cat("Recall (Sensitivity):", recall, "\n")
cat("F1-Score:", f1_score, "\n")
cat("Support:", support, "\n")
```
<br/><br/>
<h2><b>KNN Classifier</b></h2>
<br/><br/>
<h3>Training and testing model</h3>
```{r}
library(class)
k<-3
knn_predictions <- knn(train = train[, 1:9], test = test[, 1:9], cl = train$Adherence, k = k)
# Predict k-NN values on test data
# Calculate the confusion matrix
confusion_matrix <- table(knn_predictions, test$Adherence)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
cat("Accuracy:", accuracy, "\n")
```
<br/><br/>
<h3>Confusion Matrix</h3>
```{r}
# Calculate the confusion matrix and classification report
library(caret)
confusion_matrix <- confusionMatrix(as.factor(knn_predictions),as.factor(test$Adherence))

# Print the classification report
print(confusion_matrix)
```
<br/><br/>
<h3>Evaluation Metric</h3>
```{r}
# Extract precision, recall, F1-score, and support from the confusion matrix
precision <- confusion_matrix$byClass["Pos Pred Value"]
recall <- confusion_matrix$byClass["Sensitivity"]
f1_score <- confusion_matrix$byClass["F1"]
support <- confusion_matrix$byClass["Support"]

# Print the metrics
cat("Precision:", precision, "\n")
cat("Recall (Sensitivity):", recall, "\n")
cat("F1-Score:", f1_score, "\n")
cat("Support:", support, "\n")
```
<br/><br/>
<h2><b>Decision Tree</b></h2>
<br/><br/>
<h3>Training and testing model</h3>

```{r}
# Load the rpart library
library(rpart)

# Assuming you have a data frame 'data' with the target variable 'Adherence' and features

# Create and train the decision tree classifier
decision_tree_model <- rpart(Adherence ~ ., data = train, method = "class")

# Make predictions on the test data
tree_predictions <- predict(decision_tree_model, newdata = test, type = "class")

# Calculate the confusion matrix
confusion_matrix <- table(tree_predictions, test$Adherence)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
cat("Accuracy:", accuracy, "\n")
```
<br/><br/>
<h3>Confusion Matrix</h3>
```{r}
# Calculate the confusion matrix and classification report
library(caret)
confusion_matrix <- confusionMatrix(as.factor(tree_predictions),as.factor(test$Adherence))

# Print the classification report
print(confusion_matrix)
```
<br/><br/>
<h3>Evaluation Metric</h3>
```{r}
# Extract precision, recall, F1-score, and support from the confusion matrix
precision <- confusion_matrix$byClass["Pos Pred Value"]
recall <- confusion_matrix$byClass["Sensitivity"]
f1_score <- confusion_matrix$byClass["F1"]
support <- confusion_matrix$byClass["Support"]

# Print the metrics
cat("Precision:", precision, "\n")
cat("Recall (Sensitivity):", recall, "\n")
cat("F1-Score:", f1_score, "\n")
cat("Support:", support, "\n")
```
<h1><b>Comparing model performance<b><h1>
<br/><br/>
<h3>ROC curve<h3>
```{r}
library(pROC)

# List of model predictions
model_predictions <- list(
  as.numeric(log_predicted_labels),
  as.numeric(knn_predictions),
  as.numeric(tree_predictions)
  # Add more models if needed
)

# Labels for models (for the legend)
model_labels <- c("Logistic Regression", "KNN", "Decision Tree")

# Initialize an empty plot for the ROC curve
plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1), xlab = "False Positive Rate", ylab = "True Positive Rate", main = "Consolidated ROC Curve")

# Loop through each model's predictions and plot the ROC curve
colors <- 1:length(model_predictions)
for (i in 1:length(model_predictions)) {
  roc_curve <- roc(test$Adherence, model_predictions[[i]])
  lines(roc_curve, col = colors[i])
}

# Add a legend for model labels
legend("bottomright", legend = model_labels, col = colors, lty = 1)


```
<br/><br/>
<h3>AUC values of models<h3>
```{r}
# Calculate and print AUC for each model
auc_values <- sapply(model_predictions, function(pred) {
  auc(roc(test$Adherence, pred))
})

for (i in 1:length(auc_values)) {
  cat("AUC for model ", model_labels[i], ": ", auc_values[i],"\n")
}
```



